The Key to a Performant Data Parallel Compiler

### 2016-11-16 2017-spring-pldi/hsu-2017-key_data_parallel_compiler.tex
Clean up the body, intro, and notation

\begin{abstract}
We present a language-driven strategy for the construction of compilers that are inherently data-parallel in their design and implementation. Using an encoding of the inter-node relationships between nodes in an AST called a Node Co-ordinate Matrix, we demonstrate how an operator called the Key operator, that applies a function over groupings of array cells grouped and ordered by their keys, when used in conjunction with a Node Coordinate Matrix, permits arbi-trary computation over sub-trees of an AST using purely data-parallel array programming techniques. We discuss the use of these techniques when applied to an experi-mental commercial compiler called Co-dfns, which is a fully data-parallel compiler developed using the techniques discussed here.
\end{abstract}

@@

Compilers represent a peculiar class of tree/graph algo-rithms that greatly alter the underlying structure of the input graph into something very different in structure, but equivalent by some measure of interpretation. These algo-rithms are widely applicable, but have stubbornly resisted a general approach to parallelization. Most compilers are single-threaded or make very limited use of parallelism in an ad hoc way. The design and analysis of compilers al-most universally deals with compilers as recursive travers-als over ASTs. Such formulations usually include heavy reliance on branching, recursion, and sometimes very so-phisticated control flow in practice. All of this contributes to make it difficult to effectively parallelize such algo-rithms onto architectures that are sensitive to branching and recursion, such as GPGPUs or highly vectorised CPUs.

@@

We present a language-driven strategy for creating an inherently parallel compiler. We select a set of well known data-parallel primitives, and then restrict program construction to the composition of these operations into functions, without any forms of non-linear control flow, such as branching, recursion, or pattern matching. Programs constructed in such a style are data-parallel by construction. A key problem to the construction of compilers in such a restricted environment is dealing with the inherently recursive and nested structure of the AST. By working over a linearized representation of the AST as a matrix, combined with a structure we call a Node Coordinate Matrix, and using data-parallel primitives, we are able to tame the recursive and nested nature of the AST into something that can be processed handily using only data-parallel, data-flow programs, without requiring branching or other forms of non-data parallel control flow. The methods are general and independent of the language being compiled.

In the following sections we describe the core methods and idioms surrounding the AST encoding, construction of the Node Coordinate Matrix, and the use of these structures in handling nested, recursive AST relationships. This enables us to perform arbitrary computation over arbitrary sub-trees of an AST selected by parent-child relationships in the manner often seen in compiler transformations. These ideas have been further implemented and tested through the implementation of a complete data-parallel by construction compiler, called Co-dfns, that compiles a lexically scoped, functionally oriented dialect of APL with nested functions \cite{hsu2014co,hsu2015accelerating}. The compiler targets both the GPU and the CPU, and its core is implemented in this pure, restricted language that uses only function composition over data-parallel primitives.

@@

\item We describe a method of computing over arbitrary sub-trees selected by their parent-child relationships in a data-parallel manner using the Key operator and Node Coordinates.

\item We situation this technique into a broader, language-driven strategy for compiler construction that enables the development of parallel by construction compilers that are high-level in their implementation, exceptionally concise, and independent of the language being compiled.

@@

@@

\noindent That is, we indent the expression by 6 spaces, followed by the value of the expression without indent. All expressions are evaluated right to left and all function application is infix; that is, a function application may apply a function (such as \verb;+;) to one or two arguments, called monadic and dyadic application respectively, which must appear on the right and (optionally) on the left of the function name, as in the above example. Tables \ref{tab:scalarprims} to \ref{tab:conjunctions} provide a listing of all of the primitives used in the Co-dfns compiler, a selection of which are used here.

We separate out higher-order functions called operators, from functions that operate over arrays that we call simply functions. Operators take a function argument or two, and bind more strongly to the left than the right. An operator may take one or two operands, either on the left, or on the right and left, and will return a function. In the following example, we apply the reduction operator (\verb;/;) to a function created using the composition operator (\verb;∘;), as an example:

@@

\noindent We introduce notation as needed throughout the exposition, so the reader is encouraged to refer to the tables as necessary to recall particular operations.

@@

\begin{verbatim}
      (1↓Fk){⊂(0 'F' 0 ⍺)⍪⍵}⌸1↓Fast
┌─────────────────┬─────────────────┐
│┌─┬─┬─┬─────────┐│┌─┬─┬─┬─────────┐│
││0│F│0│1 0 0 0 0│││0│F│0│1 1 1 0 0││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││1│E│0│1 1 0 0 0│││3│E│0│1 1 1 1 0││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││2│F│0│1 1 1 0 0│││4│V│⍵│1 1 1 1 1││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││2│A│0│1 1 2 0 0│││4│P│+│1 1 1 1 2││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││3│N│7│1 1 2 2 0│││4│V│⍵│1 1 1 1 3││
│└─┴─┴─┴─────────┘│└─┴─┴─┴─────────┘│
└─────────────────┴─────────────────┘ 
      hd←0 'F' 0,∘⊂⊣      ⍝ Func. Header
      dp←(1+⊢-⊃)0⌷⍤1⊢     ⍝ New Depths
      rf←'V'@('F'∘=)1⌷⍤1⊢ ⍝ F → V
      (1↓Fk)(⊂hd⍪dp,rf,2↓⍤1⊢)⌸1↓Fast
┌─────────────────┬─────────────────┐
│┌─┬─┬─┬─────────┐│┌─┬─┬─┬─────────┐│
││0│F│0│1 0 0 0 0│││0│F│0│1 1 1 0 0││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││1│E│0│1 1 0 0 0│││1│E│0│1 1 1 1 0││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││2│V│0│1 1 1 0 0│││2│V│⍵│1 1 1 1 1││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││2│A│0│1 1 2 0 0│││2│P│+│1 1 1 1 2││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││3│N│7│1 1 2 2 0│││2│V│⍵│1 1 1 1 3││
│└─┴─┴─┴─────────┘│└─┴─┴─┴─────────┘│
└─────────────────┴─────────────────┘
\end{verbatim}

@@

The final task to complete function lifting of each function body before lifting is to shift the depths in the depth vectors to correspond to those of a function lifted to the top-level and to attach a function node to the top of each of the bodies. At that point, we simply recombine all of the newly created functions into a single top level.

@@

@@

\begin{verbatim}
      (1↓Ek){⍺ ⍵}⌸1↓East
┌─────┬─────────────┐
│1 0 0│┌─┬─┬─┬─────┐│
│     ││1│E│0│1 1 0││
│     │├─┼─┼─┼─────┤│
│     ││1│P│÷│1 2 0││
│     │├─┼─┼─┼─────┤│
│     ││1│E│0│1 3 0││
│     │└─┴─┴─┴─────┘│
├─────┼─────────────┤
│1 1 0│┌─┬─┬─┬─────┐│
│     ││2│V│a│1 1 1││
│     │├─┼─┼─┼─────┤│
│     ││2│P│+│1 1 2││
│     │├─┼─┼─┼─────┤│
│     ││2│V│b│1 1 3││
│     │└─┴─┴─┴─────┘│
├─────┼─────────────┤
│1 3 0│┌─┬─┬─┬─────┐│
│     ││2│V│c│1 3 4││
│     │├─┼─┼─┼─────┤│
│     ││2│P│×│1 3 5││
│     │├─┼─┼─┼─────┤│
│     ││2│V│d│1 3 6││
│     │└─┴─┴─┴─────┘│
└─────┴─────────────┘
      hd←0'E'0,∘⊂⊣
      rf←'V'@('E'∘=)1⌷⍤1⊢
      ⊖(1↓Ek)(hd⍪dp,rf,2↓⍤1⊢)⌸1↓East
┌─┬─┬─┬─────┐
│0│E│0│1 3 0│
├─┼─┼─┼─────┤
│1│V│c│1 3 4│
├─┼─┼─┼─────┤
│1│P│×│1 3 5│
├─┼─┼─┼─────┤
│1│V│d│1 3 6│
└─┴─┴─┴─────┘
┌─┬─┬─┬─────┐
│0│E│0│1 1 0│
├─┼─┼─┼─────┤
│1│V│a│1 1 1│
├─┼─┼─┼─────┤
│1│P│+│1 1 2│
├─┼─┼─┼─────┤
│1│V│b│1 1 3│
└─┴─┴─┴─────┘
┌─┬─┬─┬─────┐
│0│E│0│1 0 0│
├─┼─┼─┼─────┤
│1│V│0│1 1 0│
├─┼─┼─┼─────┤
│1│P│÷│1 2 0│
├─┼─┼─┼─────┤
│1│V│0│1 3 0│
└─┴─┴─┴─────┘
\end{verbatim}

Again, we can see immediately that we have grouped each set of nodes according to the expressions that are to be lifted. Just as in the case of function lifting, we can adjust the depths of each expression to the correct depth and we can replace each expression node with a variable reference based on that node’s coordinate. Each expression can be given a unique name based on its coordinate. A later compiler pass can reduce these names down to the minimum actually required to represent the expression if desired.

The only extra issue involved here is to ensure that the order of evaluation matches. In our case, we are assuming that the order of evaluation is right to left, which means that the above order is actually backwards of the desired order. During recombination, we simply reverse these orders and this fixes that problem. More work would be required to take into consideration a specific precedence hierarchy.

@@

>>>>

\begin{abstract}
We present a language-driven strategy for the construction of compilers that are inherently data-parallel in their design and implementation. Using an encoding of the inter-node relationships between nodes in an AST called a Path Matrix, we demonstrate how an operator called the Key operator, that applies a function over groupings of array cells grouped and ordered by their keys, when used in conjunction with a Path Matrix, permits arbitrary computation over sub-trees of an AST using purely data-parallel array programming techniques. We discuss the use of these techniques when applied to an experimental commercial compiler called Co-dfns, a fully data-parallel compiler developed using the techniques discussed here. On transformation heavy operations such as expression flattening or function lifting, these techniques, when compared against pure and effectful recursive implementations, outperforms the CPU-based algorithms by a factor of 10 - 58 on the GPU. 
\end{abstract}

@@

Compilers represent a peculiar class of tree/graph algorithms that greatly alter the underlying structure of the input graph into something very different in structure, but equivalent by some semantic interpretation. These algorithms are widely applicable, but have stubbornly resisted a general approach to parallelization. Most compilers are single-threaded or make very limited use of parallelism in an ad hoc way. The design and analysis of compilers almost universally deals with compilers as recursive traversals over ASTs. Such formulations usually include heavy reliance on branching, recursion, and sometimes very sophisticated control flow. All of this contributes to make it difficult to effectively parallelize such algo-rithms onto architectures that are sensitive to branching and recursion, such as GPGPUs or highly vectorised CPUs.

@@

We present a language-driven strategy for creating an inherently parallel compiler. We select a set of well known data-parallel primitives, and then restrict program construction to the composition of these operations into functions, without any forms of non-linear control flow, such as branching, recursion, or pattern matching. Programs constructed in such a style are data-parallel by construction. A key problem to the construction of compilers in such a restricted environment is dealing with the inherently recursive and nested structure of the AST. By working over a linearized representation of the AST as a matrix, combined with a structure we call a Path Matrix, and using data-parallel primitives, we are able to tame the recursive and nested nature of the AST into something that can be processed handily using only data-parallel, data-flow programs, without requiring branching or other forms of non-data parallel control flow. The methods are general and independent of the language being compiled.

In the following sections we describe the core methods and idioms surrounding the AST encoding, construction of the Path Matrix, and the use of these structures in handling nested, recursive AST relationships. This enables us to perform arbitrary computation over arbitrary sub-trees of an AST selected by parent-child relationships in the manner often seen in compiler transformations. These ideas have been further implemented and tested through the implementation of a complete data-parallel by construction compiler, called Co-dfns, that compiles a lexically scoped, functionally oriented dialect of APL with nested functions \cite{hsu2014co,hsu2015accelerating}. The compiler targets both the GPU and the CPU, and its core is implemented in this pure, restricted language that uses only function composition over data-parallel primitives.

To illustrate the performance of these techniques, we micro-benchmark a transformation heavy expression lifting algorithm that is found almost universally in compilers. This benchmark shows performance speedups of up to a factor of 58 when compared against pure, functionally oriented algorithms.

@@

\item We describe a method of computing over arbitrary sub-trees selected by their parent-child relationships in a data-parallel manner using the Key operator and a Path Matrix.

\item We situate this technique into a broader, language-driven strategy for compiler construction that enables the development of parallel by construction compilers that are high-level in their implementation, exceptionally concise, and independent of the language being compiled.

@@

\item We compare the performance of the Key-based algorithms against traditional recursive methods for the task of lifting/flattening expressions, such as commonly done for function lifting and expression flattening. Our results demonstrate consistent speedups over the traditional methods.

@@

\noindent That is, we indent the expression by 6 spaces, followed by the value of the expression without indent. All expressions are evaluated right to left and all function application is infix; that is, a function application may apply a function (such as \verb;+;) to one or two arguments, called monadic and dyadic application respectively, which must appear on the right and (optionally) on the left of the function name, as in the above example. The Appendix provides a listing of all of the primitives used in the Co-dfns compiler, a selection of which are used here.

We differentiate higher-order functions called operators, from functions that operate over arrays that we call simply functions. Operators take a function argument or two, and bind more strongly to the left than the right. An operator may take one or two operands, either on the left, or on the right and left, and will return a function. In the following example, we apply the reduction operator (\verb;/;) to a function created using the composition operator (\verb;∘;), as an example:

@@

\noindent We introduce notation as needed throughout the exposition, so the reader is encouraged to refer to the appendix as necessary to recall particular operations.

@@

Each of these grouped sub-trees will be given a new local root \verb;F; node that is at the appropriate depth (0) and uses the key as its path vector (\verb;⊣;).

\begin{verbatim}
      (1↓Fk){⊂(0 'F' 0 ⍺)⍪⍵}⌸1↓Fast
┌─────────────────┬─────────────────┐
│┌─┬─┬─┬─────────┐│┌─┬─┬─┬─────────┐│
││0│F│0│1 0 0 0 0│││0│F│0│1 1 1 0 0││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││1│E│0│1 1 0 0 0│││3│E│0│1 1 1 1 0││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││2│F│0│1 1 1 0 0│││4│V│⍵│1 1 1 1 1││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││2│A│0│1 1 2 0 0│││4│P│+│1 1 1 1 2││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││3│N│7│1 1 2 2 0│││4│V│⍵│1 1 1 1 3││
│└─┴─┴─┴─────────┘│└─┴─┴─┴─────────┘│
└─────────────────┴─────────────────┘ 
\end{verbatim}

\noindent At this point, the internal function nodes refer by path vector to the newly added local roots. This was down without traversing or maintaining these connections, but falls out as a natural artifact of using the Key (\verb;⌸;) operator. At this point we are ready to fix up the sub-tree to be correct. Firstly, we need a function to encapsulate the function header information.

\begin{verbatim}
      hd←0 'F' 0,∘⊂⊣      ⍝ Func. Header
\end{verbatim}

\noindent Next, we write a function to adjust the depths, which is one more than the difference between each node's depth and the depth of the first node of the sub-tree.

\begin{verbatim}
      dp←(1+⊢-⊃)0⌷⍤1⊢     ⍝ New Depths
\end{verbatim}

\noindent Finally, we use the \verb;@; operator to replace occurances of \verb;F; in the Type column with \verb;V; elements.

\begin{verbatim}
      rf←'V'@('F'∘=)1⌷⍤1⊢ ⍝ F → V
\end{verbatim}

\noindent Once we put all this together, we can construct the new sub-trees with the appropriate depths, types, references, and links with the following expression.

\begin{verbatim}
      (1↓Fk)(⊂hd⍪dp,rf,2↓⍤1⊢)⌸1↓Fast
┌─────────────────┬─────────────────┐
│┌─┬─┬─┬─────────┐│┌─┬─┬─┬─────────┐│
││0│F│0│1 0 0 0 0│││0│F│0│1 1 1 0 0││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││1│E│0│1 1 0 0 0│││1│E│0│1 1 1 1 0││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││2│V│0│1 1 1 0 0│││2│V│⍵│1 1 1 1 1││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││2│A│0│1 1 2 0 0│││2│P│+│1 1 1 1 2││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││3│N│7│1 1 2 2 0│││2│V│⍵│1 1 1 1 3││
│└─┴─┴─┴─────────┘│└─┴─┴─┴─────────┘│
└─────────────────┴─────────────────┘
\end{verbatim}

@@

@@

Flattening expressions is almost identical to lifting functions, in part because we have delayed the lexical scope resolution aspect of lifting functions to a later pass that will handle this explicitly. In addition to lifting the expressions themselves, however, we will also take the opportunity to handle precedence and ordering. In our examples, we are assuming that all expressions evaluate right to left. Because the Key (\verb;⌸;) operator is a stable operation, and we require that nodes have a depth-first pre-order traversal pattern, then we can trivially handle precedence order of this type, including handling parentheses, by reversing the results of flattening.

@@

\begin{verbatim}
      (1↓Ek){⍺ ⍵}⌸1↓East
┌─────┬─────────────┐
│1 0 0│┌─┬─┬─┬─────┐│
│     ││1│E│0│1 1 0││
│     │├─┼─┼─┼─────┤│
│     ││1│P│÷│1 2 0││
│     │├─┼─┼─┼─────┤│
│     ││1│E│0│1 3 0││
│     │└─┴─┴─┴─────┘│
├─────┼─────────────┤
│1 1 0│┌─┬─┬─┬─────┐│
│     ││2│V│a│1 1 1││
│     │├─┼─┼─┼─────┤│
│     ││2│P│+│1 1 2││
│     │├─┼─┼─┼─────┤│
│     ││2│V│b│1 1 3││
│     │└─┴─┴─┴─────┘│
├─────┼─────────────┤
│1 3 0│┌─┬─┬─┬─────┐│
│     ││2│V│c│1 3 4││
│     │├─┼─┼─┼─────┤│
│     ││2│P│×│1 3 5││
│     │├─┼─┼─┼─────┤│
│     ││2│V│d│1 3 6││
│     │└─┴─┴─┴─────┘│
└─────┴─────────────┘
\end{verbatim}

\noindent Again, we can see immediately that we have grouped each set of nodes according to the expressions that are to be lifted. Just as in the case of function lifting, we can adjust the depths of each expression to the correct depth and we can replace each expression node with a variable reference based on that node’s coordinate. Each expression can be given a unique name based on its coordinate. A later compiler pass can reduce these names down to the minimum actually required to represent the expression if desired. As with the function lifting case, we need some functions to help abstract each new field. In our case, we use the same functions as with function lifting but alter the \verb;hd; and \verb;rf; functions to use the right node types:

\begin{verbatim}
      hd←0'E'0,∘⊂⊣
      rf←'V'@('E'∘=)1⌷⍤1⊢
\end{verbatim}

\noindent At this point, the expression lifting operation is then almost identical to that of function lifting. Notice that we reverse the 3-D matrix result at the end along the first access to ensure that order of evaluation is preserved correctly.

\begin{verbatim}
      ⊖(1↓Ek)(hd⍪dp,rf,2↓⍤1⊢)⌸1↓East
┌─┬─┬─┬─────┐
│0│E│0│1 3 0│
├─┼─┼─┼─────┤
│1│V│c│1 3 4│
├─┼─┼─┼─────┤
│1│P│×│1 3 5│
├─┼─┼─┼─────┤
│1│V│d│1 3 6│
└─┴─┴─┴─────┘
┌─┬─┬─┬─────┐
│0│E│0│1 1 0│
├─┼─┼─┼─────┤
│1│V│a│1 1 1│
├─┼─┼─┼─────┤
│1│P│+│1 1 2│
├─┼─┼─┼─────┤
│1│V│b│1 1 3│
└─┴─┴─┴─────┘
┌─┬─┬─┬─────┐
│0│E│0│1 0 0│
├─┼─┼─┼─────┤
│1│V│0│1 1 0│
├─┼─┼─┼─────┤
│1│P│÷│1 2 0│
├─┼─┼─┼─────┤
│1│V│0│1 3 0│
└─┴─┴─┴─────┘
\end{verbatim}

The only extra issue involved here over function lifting is to ensure that the order of evaluation matches. In our case, we are assuming that the order of evaluation is right to left, which means that the above order given by \verb;⌸; is actually backwards of the desired order. During recombination, we simply reverse these orders and this fixes that problem. More work would be required to take into consideration a specific precedence hierarchy.

@@

It should also be noted that while the mutation-oriented algorithm is obviously faster than the functional algorithm on the CPU, it comes at the cost of stylistic clarity and the capability to reason locally about computation. Our Key-based algorithm remains pure and functional while still being fast. It is also exceptionally concise relative to the other two algorithms.
